---
title: "single_cell_subset_DE_GO_analysis"
author: "Johannes Schöneich"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
# The idea of this script is to do a DE/GO analysis on single cell
# on the subsets of Absorptive Enterocytes and Goblet/Paneth
library(circlize)
library(clusterProfiler)
library(DOSE)
library(EnhancedVolcano)
library(enrichplot)
library(ComplexHeatmap)
library(glue)
library(openxlsx)
library(patchwork)
library(RColorBrewer)
library(Scillus)
library(scCustomize)
library(Seurat)
library(SeuratWrappers)
library(stringr)
library(tidyverse)

harddrive <- "H:/231248/NTFS_3.64TB"

source(file.path(harddrive, "scrna_datasets/save_load_helper.R"))
source(file.path(harddrive, "scrna_datasets/single_cell_themes.R"))
source(file.path(harddrive, "helper_functions.R"))
source(file.path(harddrive, "DE_analysis/DE_helperfunctions.R"))

analysis_folder <- file.path(harddrive, "subsets/subsets_2025/DE_GO")
dir.create(analysis_folder, recursive = TRUE)
```

```{r}
# Define the folder path for Enterocyte datasets
ent_folder <- file.path(
  harddrive,
  "subsets/subsets_2025/save/noc2_no11/ent"  
)

# Load Enterocyte dataset
scrna_ent_inf <- load_object(file.path(
  ent_folder, 
  "scrna_ent_inf_pseudotime.Rds" 
))

# Define the folder path for Goblet datasets
goblet_folder <- file.path(
  harddrive,
  "subsets/subsets_2025/save/noc2/gob/"  
)

scrna_gob_inf <- load_object(file.path(
  goblet_folder, 
  "scrna_phase_clustering.Rds" 
))
```

```{r}
Idents(scrna_ent_inf) <- "stage"
scrna_ent_uninf <- subset(scrna_ent_inf,idents = "d05i",invert = TRUE)

Idents(scrna_gob_inf) <- "stage"
scrna_gob_uninf <- subset(scrna_gob_inf,idents = "d05i",invert = TRUE)
```

```{r}
save_object(scrna_ent_uninf,file.path(ent_folder,"scrna_ent_uninf.Rds"))
save_object(scrna_gob_uninf,file.path(goblet_folder,"scrna_gob_uninf.Rds"))
```


```{r}
Idents(scrna_gob_uninf) <- "stage"
results_list_goblet <- calculate_results_list_presto(scrna_gob_uninf)

Idents(scrna_ent_uninf) <- "stage"
results_list_ent <- calculate_results_list_presto(scrna_ent_uninf)
```

```{r}
plots_folder <- file.path(analysis_folder,"DE/plots")
charts_folder <- file.path(analysis_folder,"DE/charts")

GO_save_folder <- file.path(analysis_folder,"GO/saves")

# Which ontology: choose between all, biological process (bp), 
# molecular function (mf) and cellular component (cc).
ont <- "bp"

# P-adjust cutoff. Adjustment method is benjamini-hochberg (FDR)
p.cutoff <- 0.05

# The number of GO terms to compare per cluster
topN  <- 3

# Count cutoff. The low end filtering of counts
count.cutoff <- 10

# Heatmap coloring options
pos_color_low <- "#fff5f0"
pos_color_high <- "#E41A1C" # The color to which the significant p-values converge

neg_color_low <- "#F7FBFF"
neg_color_high <- "#377EB8"

# One of GO, KEGG, Reactome or all
comparison <- "GO"

#analysis name for plot names
NAME <- "GO_goblet"

dir.create(plots_folder,recursive = TRUE)
dir.create(charts_folder,recursive = TRUE)
dir.create(GO_save_folder,recursive = TRUE)
```

## Volcano Plots DE genes

The contrast order is always +logFC then -logFC 
(first numerator (Zähler) then Denominator (Nenner))
condition treated vs untreated, tells you that the estimates are of the 
logarithmic fold change log2(treated/untreated).

The level given last is the base level for the comparison. The syntax is:
contrast <- c("condition", "level_to_compare", "base_level")
for example, if we observe a log2 fold change of -2 this would mean 
the gene expression is lower in the level of interest relative to the base level

```{r fig.height = 10, fig.width = 10}
generate_volcano_plots(results_list = results_list_goblet,
                       pos_color = "#E41A1C",
                       neg_color = "#377EB8",
                       plots_folder = plots_folder,
                       charts_folder = charts_folder,
                       filename_prefix = "single_cell_goblet",
                       do_save = TRUE)

# we only need consecutive timepoints
results_list_goblet <- results_list_goblet[c(1,4,6)]

generate_volcano_plots(results_list = results_list_ent,
                       pos_color = "#E41A1C",
                       neg_color = "#377EB8",
                       plots_folder = plots_folder,
                       charts_folder = charts_folder,
                       filename_prefix = "single_cell_enterocyte",
                       do_save = TRUE)

# we only need consecutive timepoints
results_list_ent <- results_list_ent[c(1,4,6)]
```

# Calculate GO/Pathways and compare cluster
```{r}
res_up <- lapply(results_list_ent,function(df){
  sigsdf_up <- df[(df$avg_log2FC >= 1) & (df$p_val_adj <= 0.05),]
  upgene <- (sigsdf_up$symbol)
  return(upgene)
})

res_down <- lapply(results_list_ent,function(df){
  sigsdf_down <- df[(df$avg_log2FC <= - 1) & (df$p_val_adj <= 0.05),]
  downgene <- (sigsdf_down$symbol)
  return(downgene)
})

up_compare_GO <- compareCluster(res_up,
                                fun = enrichGO,
                                OrgDb = "org.Mm.eg.db",
                                keyType = "SYMBOL",
                                ont = ont,
                                pAdjustMethod = "BH",
                                pvalueCutoff  = p.cutoff,
                                readable = TRUE
                             )
  
write.csv(as.data.frame(up_compare_GO),
            file = file.path(charts_folder,
                             glue("up_compare_{NAME}.csv")),
            row.names = FALSE)

down_compare_GO <- compareCluster(res_down,
  fun = enrichGO,
  OrgDb = "org.Mm.eg.db",
  keyType = "SYMBOL",
  ont = ont,
  pAdjustMethod = "BH",
  pvalueCutoff  = p.cutoff,
  readable = TRUE
  )

write.csv(as.data.frame(down_compare_GO),
            file = file.path(charts_folder,
                             glue("down_compare_{NAME}.csv")),
            row.names = FALSE)
```

# Save results
```{r}
saveRDS(down_compare_GO,
        file.path(GO_save_folder,glue("down_compare_{NAME}.Rds")))

saveRDS(up_compare_GO,
        file.path(GO_save_folder,glue("up_compare_{NAME}.Rds")))
```

```{r}
down_compare_GO <- readRDS(file.path(GO_save_folder,glue("down_compare_{NAME}.Rds")))

up_compare_GO <- readRDS(file.path(GO_save_folder,glue("up_compare_{NAME}.Rds")))
```


# Calculations
```{r}
# Filter and arrange by p-value
up_compare_filter <- filter(up_compare_GO,
                            Count > count.cutoff) %>% arrange(p.adjust)

# -log of adjusted p.value
up_compare_filter@compareClusterResult$p.adjust <- 
  -log10(up_compare_filter@compareClusterResult$p.adjust)

# Filter and arrange by p.value
down_compare_filter <- filter(down_compare_GO,
                              Count > count.cutoff) %>% arrange(p.adjust)

# -log of p.value
down_compare_filter@compareClusterResult$p.adjust <- -log10(
  down_compare_filter@compareClusterResult$p.adjust )
```

# Maximal and minimal p-values for scales
```{r}
max_p <- max(down_compare_filter@compareClusterResult$p.adjust,
                     up_compare_filter@compareClusterResult$p.adjust)

if(max_p > 30){max_p <- 30}
```

# Fortify data.frames for plotting (topN)
```{r}
# Make data frames for plotting. Only use topN terms per Cluster
df_up <- fortify(up_compare_filter, 
                         showCategory = topN, split = NULL)
df_up <- droplevels(df_up)

df_down <- fortify(down_compare_filter, 
                           showCategory = topN, split = NULL)
df_down <- droplevels(df_down)

x_labels_up <- gsub("\n"," ",levels(df_up$Cluster))
x_labels_down <- gsub("\n"," ",levels(df_down$Cluster))

levels(df_up$Cluster) <- x_labels_up
levels(df_down$Cluster) <- x_labels_down
```

# Clustering
```{r}
clust_met <- "complete"

markers.down <- df_down$Description %>% unique()
markers.up <- df_up$Description %>% unique()

p.mat.down <- df_down %>% 
  filter(Description %in% markers.down) %>% 
  dplyr::select(Description,Cluster,p.adjust) %>%
  pivot_wider(names_from = Cluster,values_from = p.adjust) %>%
  data.frame() # make df
row.names(p.mat.down) <- p.mat.down$Description  # put gene in `row`
p.mat.down <- p.mat.down[,-1] #drop gene column as now in rows
p.mat.down[is.na(p.mat.down)] <- 0

# hclust with distance matrix
clust.down <- hclust(dist(p.mat.down %>% as.matrix()),method = clust_met) 

p.mat.up <- df_up %>% 
  filter(Description %in% markers.up) %>% 
  dplyr::select(Description,Cluster,p.adjust) %>%
  pivot_wider(names_from = Cluster,values_from = p.adjust) %>%
  data.frame() # make df as tibbles -> matrix annoying
row.names(p.mat.up) <- p.mat.up$Description  # put gene in `row`
p.mat.up <- p.mat.up[,-1] #drop gene column as now in rows
p.mat.up[is.na(p.mat.up)] <- 0

# hclust with distance matrix
clust.up <- hclust(dist(p.mat.up %>% as.matrix()),method = clust_met) 
```

# Plotting
```{r}
theme.dp <- list(
  geom_point(),
  theme_dose(7),
  #coord_fixed(ratio = 1.4),
  scale_color_gradient(low = pos_color_low,
                       high = pos_color_high,
                       limits = c(0,max_p)), #adjust limits
  ylab(NULL),
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 90,
                               hjust = 1,
                               vjust = .5),
    axis.text.y = element_text(angle = 25,
                               vjust = 0.5,
                               margin = margin(r = -10) ),
        legend.text = element_text(size = 7),
        legend.position = "inside",
        legend.position.inside =  c(2.7,-0.1), # GO
        #legend.position = c(3,-0.1), # kegg
        #legend.position = c(2.5,-0.1), # reactome
        legend.direction = "vertical",
        legend.box = "vertical",
        legend.spacing.x = unit(0.05, 'cm'),
        legend.title = element_text(size = 7),
    legend.title.position = "top",
        panel.background = element_rect(fill='transparent'), #transparent panel bg
    plot.background = element_rect(fill='transparent', color = NA), #transparent plot bg
   # panel.grid.major = element_blank(), #remove major gridlines
   # panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_blank(), #transparent legend panel
   axis.ticks = element_blank()
  ),
  labs(color = "-log10 (p.adjust)"),
  guides(size = guide_legend(order = 2),
         color = guide_colorbar(order = 1,
                                direction = "vertical",
                                title.theme =
                                  element_text(
                                    margin = margin(0, 0, 0, 0, "pt"))
                                )
         )
)

# GO + Pathway Up
df_up_rc <- df_up

df_up_rc$Description <- fct_relevel(df_up_rc$Description,
                                rev(clust.up$labels[clust.up$order]))

df_up_rc$Description_trunc <- sapply(df_up_rc$Description,
                                     truncate_go_term, max_length = 30)

df_up_rc$ID_desc <- paste(df_up_rc$ID,df_up_rc$Description_trunc)

df_up_rc <- df_up_rc %>%
  mutate(p.adjust = ifelse(p.adjust > max_p, max_p, p.adjust))

df_up_rc <- mutate(df_up_rc, 
                     FoldEnrichment = GeneRatio / parse_ratio(BgRatio))

df_up_rc <- 
  mutate(df_up_rc,
    ClusterGenes = as.numeric(gsub(".*\\((\\d+)\\).*", "\\1", Cluster)),
    Norm_GR = GeneRatio / ClusterGenes
  )

# GO + Pathway Down
df_down_rc <- df_down

df_down_rc$Description <- fct_relevel(df_down$Description,
                                rev(clust.down$labels[clust.down$order]))

df_down_rc$Description_trunc <- sapply(df_down_rc$Description,
                                       truncate_go_term,max_length = 30)

df_down_rc$ID_desc <- paste(df_down_rc$ID,df_down_rc$Description_trunc)

df_down_rc <- df_down_rc %>%
  mutate(p.adjust = ifelse(p.adjust > max_p, max_p, p.adjust))

df_down_rc <- mutate(df_down_rc, 
                     FoldEnrichment = GeneRatio / parse_ratio(BgRatio))

df_down_rc <- mutate(df_down_rc,
    ClusterGenes = as.numeric(gsub(".*\\((\\d+)\\).*", "\\1", Cluster)),
    Norm_GR = GeneRatio / ClusterGenes
  )
```

```{r}
dp_up_FE <- ggplot(df_up_rc, aes(x = Cluster, 
                                 y = Description, 
                                 size = FoldEnrichment,
                                 color = p.adjust,
                                 stroke = NA)) +
  theme.dp + scale_y_discrete(position = "right")

dp_down_FE <- ggplot(df_down_rc, aes(x = Cluster,
                                     y = Description, 
                                     size = FoldEnrichment,
                                     color = p.adjust, 
                                     stroke = NA)) +
  theme.dp + scale_y_discrete(position = "right") +
  scale_color_gradient(low = neg_color_low,
                       high = neg_color_high,
                       limits = c(0,max_p))
```

# Saving

## Save Plots
```{r}
save_ggplot_formats(
  plt = dp_up_FE,
  create_plot_subdir = TRUE,
  base_plot_dir = plots_folder,
  plt_name = glue("{NAME}_dotplot_up_rc_foldenrichment_top{topN}_",
  "count_cutoff_{count.cutoff}_v2"),
  width = 2.7,height = 3
  )

save_ggplot_formats(
  plt = dp_down_FE,
  create_plot_subdir = TRUE,
  base_plot_dir = plots_folder,
  plt_name = glue("{NAME}_dotplot_down_rc_foldenrichment_top{topN}_",
  "count_cutoff_{count.cutoff}_v2"),
  width = 2.5,height = 2
  )
```


## Save Tables
```{r}
write.csv(as.data.frame(df_down_rc),
            file = file.path(charts_folder,
                             glue("down_compare_filter_{NAME}.csv")),
            row.names = FALSE)

write.csv(as.data.frame(df_up_rc),
            file = file.path(charts_folder,
                             glue("up_compare_filter_{NAME}.csv")),
            row.names = FALSE)
```



```{r}
dp1 <- DimPlot(scrna_ent_uninf,reduction = "INTE_UMAP",group.by = "stage")
dp2 <- DimPlot(scrna_ent_uninf,reduction = "INTE_UMAP",group.by = "integrated_snn_res.0.8")

save_ggplot_formats(
  plt = dp1,
  create_plot_subdir = TRUE,
  base_plot_dir = plots_folder,
  plt_name = glue("stage_plot"),
  width = 4,height = 4
  )

save_ggplot_formats(
  plt = dp2,
  create_plot_subdir = TRUE,
  base_plot_dir = plots_folder,
  plt_name = glue("int_plot"),
  width = 4,height = 4
  )
```

