---
title: "DE_scRNA"
author: "Johannes Schöneich"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(EnhancedVolcano)
library(tidyverse)
library(Seurat)
library(limma)
library(glue)
library(VennDiagram)
library(cowplot)
library(ComplexHeatmap)
library(clusterProfiler)
library(enrichplot)
library(DOSE)
library(limma)
library(ReactomePA)
library(openxlsx)
library(cowplot)
library(gridExtra)
library(patchwork)
library(openxlsx)

harddrive <- "H:"

source(file.path(harddrive,"save_load_helper.R"))
source(file.path(harddrive,"helper_functions.R"))
source(file.path(harddrive,"DE_analysis/DE_helperfunctions.R"))
source(file.path(harddrive,"scrna_datasets/single_cell_themes.R"))
```

```{r}
tools <- load_object(file.path(harddrive,"scrna_datasets/scrna_both_d25/",
                      "scrna_full_2024_both_d25_broad_inf/save/",
                      "tools_scrna_broad_inf.Rds"))

analysis_folder <- "scrna_datasets/infected_d05"

plots_folder <- file.path(harddrive,analysis_folder,"DE/plots")
charts_folder <- file.path(harddrive,analysis_folder,"DE/charts")

GO_save_folder <- file.path(analysis_folder,
                          "GO/saves/")

pathway_save_folder <- file.path(analysis_folder,
                          "Pathway/saves/")

# Which ontology: choose between all, biological process (bp), 
# molecular function (mf) and cellular component (cc).
ont <- "bp"

# P-adjust cutoff. Adjustment method is benjamini-hochberg (FDR)
p.cutoff <- 0.05

# The number of GO terms to compare per cluster
topN  <- 5

# Count cutoff. The low end filtering of counts
count.cutoff <- 5

# Heatmap coloring options
pos_color_low <- "#fff5f0"
pos_color_high <- "#E41A1C" # The color to which the significant p-values converge

neg_color_low <- "#F7FBFF"
neg_color_high <- "#377EB8"

# One of GO, KEGG, Reactome or all
comparison <- "GO"

#analysis name for plot names
NAME <- "GO"

dir.create(analysis_folder,recursive = TRUE)
dir.create(plots_folder,recursive = TRUE)
dir.create(charts_folder,recursive = TRUE)
dir.create(GO_save_folder,recursive = TRUE)
dir.create(pathway_save_folder,recursive = TRUE)

# read in the results.list if this analysis was already run 
# and you want to change plots
results.list <- tools[["dego_name_int_0.3_broad"]]$de$d05.vs.d05i
```

## Volcano Plots DE genes

The contrast order is always +logFC then -logFC 
(first numerator (Zähler) then Denominator (Nenner))
condition treated vs untreated, tells you that the estimates are of the 
logarithmic fold change log2(treated/untreated).

The level given last is the base level for the comparison. The syntax is:
contrast <- c("condition", "level_to_compare", "base_level")
for example, if we observe a log2 fold change of -2 this would mean 
the gene expression is lower in the level of interest relative to the base level

```{r}
#change logfc so that infected is right, uninfected left
results.list <- lapply(results.list,
                       function(res){
                         res$avg_log2FC <- -res$avg_log2FC
                         return(res)
                         }
                       )
```

```{r fig.height = 10, fig.width = 10}
generate_volcano_plots(results_list = results.list,
                       pos_color = "#E41A1C",
                       neg_color = "#377EB8",
                       plots_folder = plots_folder,
                       charts_folder = charts_folder,
                       filename_prefix = "Full")
```

```{r}
# unique DE genes for infection

infected_genes <- lapply(results.list,function(x){
  as.data.frame(x) %>% filter(avg_log2FC >= 1 & p_val_adj <= 0.05) %>% pull(gene)
})

infected_genes$Tuft <- NULL

# Create a binary presence matrix for UpSet plot
gene_names <- unique(unlist(infected_genes))

presence_matrix <- sapply(infected_genes, function(genes) gene_names %in% genes)

rownames(presence_matrix) <- gene_names

# Convert to a logical matrix for UpSet plot
presence_matrix <- as.matrix(presence_matrix)

mat <- make_comb_mat(infected_genes)

mat_low_filter <- mat[comb_size(mat) >= 10]

# Create the UpSet plot
us_plot <- UpSet(mat_low_filter,
      top_annotation = upset_top_annotation(mat_low_filter, add_numbers = TRUE),
      right_annotation = upset_right_annotation(mat_low_filter, add_numbers = TRUE),
      comb_order = rev(order(comb_size(mat_low_filter))),
      )
```

```{r}
save_ggplot_formats(
  plt = us_plot,
  base_plot_dir = plots_folder,
  plot_obj = "not_gg",
  plt_name = "upset_unique_low_end_filter",
  width = 6, height = 3
)
```

```{r}
# Define the binary string labels corresponding to the unique genes
labels <- c("1000", "0100", "0010", "0001")

# Create an empty list to store results
unique_genes_list <- list()

# Loop over the labels and store the results
for (i in seq_along(labels)) {
  unique_genes_list[[i]] <- extract_comb(mat, labels[i])
}

# Assign meaningful names to the list elements
names(unique_genes_list) <- c("abs.ent_unique", "eec_unique", "gp_unique", "stem_unique")


# Custom function to convert a number to a binary string with leading zeros
to_binary <- function(x, digits = 4) {
  binary <- intToBits(x)[1:digits]
  paste(rev(as.integer(binary)), collapse = "")
}

# Define the names for the primary sets
names_map <- list(
  "1000" = "Abs.Ent.",
  "0100" = "EEC",
  "0010" = "Goblet+Paneth",
  "0001" = "Stem"
)

# Generate the list of all combinations
all_combinations <- list()

# Loop through all binary combinations (0001 to 1111)
for (i in 1:15) {
  # Convert to 4-digit binary string
  binary_code <- to_binary(i)
  
  # Extract geneset using the binary code
  geneset <- extract_comb(mat, binary_code)
  
  # Determine the name of the set
  if (binary_code %in% names(names_map)) {
    set_name <- names_map[[binary_code]]
  } else {
    # For combinations, concatenate the names of relevant sets
    active_sets <- strsplit(binary_code, "")[[1]] == "1"
    set_name <- paste(names_map[names(names_map)[active_sets]], collapse = " + ")
  }
  
  # Add to the list
  all_combinations[[set_name]] <- geneset
}

names(all_combinations)[15] <- "Common response"
# Save the list to an Excel file
write.xlsx(all_combinations,
           file = file.path(charts_folder,"infected_all_combinations.xlsx"), 
           asTable = FALSE)
```

# GO analysis
```{r}
genes_entrez <- lapply(unique_genes_list,function(upgene){
  gene_e <- bitr(upgene, fromType="SYMBOL", toType = "ENTREZID", 
                    OrgDb = "org.Mm.eg.db")
  return(gene_e[["ENTREZID"]])
})


if(comparison == "GO"| comparison == "all"){
  unique_infected_compare_GO <- compareCluster(unique_genes_list,
                             fun = enrichGO,
                             OrgDb = "org.Mm.eg.db",
                             keyType = "SYMBOL",
                             ont = ont,
                             pAdjustMethod = "BH",
                             pvalueCutoff  = p.cutoff,
                             readable = TRUE
                             )
  
  write.csv(as.data.frame(unique_infected_compare_GO),
            file = file.path(charts_folder,
                             "unique_infected_compare_GO.csv"),
            row.names = FALSE)
  unique_infected_compare <- unique_infected_compare_GO
}

if(comparison == "Reactome"| comparison == "all"){
  unique_infected_compare_reactome <- compareCluster(genes_entrez,
                             fun = enrichPathway,
                             organism = "mouse",
                             pAdjustMethod = "BH",
                             pvalueCutoff  = p.cutoff,
                             readable = TRUE
                             )
  
  write.csv(as.data.frame(unique_infected_compare_reactome),
            file = file.path(charts_folder,
                             "unique_infected_compare_reactome.csv"),
            row.names = FALSE)
  unique_infected_compare <- unique_infected_compare_reactome
}

if(comparison == "KEGG"| comparison == "all"){
  unique_infected_compare_KEGG <- compareCluster(genes_entrez,
                             fun = enrichKEGG,
                             organism = "mmu",
                             pAdjustMethod = "BH",
                             pvalueCutoff  = p.cutoff,
                             )
  
    unique_infected_compare_KEGG <- setReadable(unique_infected_compare_KEGG, 
                                   "org.Mm.eg.db",
                                   keyType = "ENTREZID")
    
    unique_infected_compare_KEGG@compareClusterResult$Description <- gsub(
      " - Mus musculus \\(house mouse)","",
      unique_infected_compare_KEGG@compareClusterResult$Description)
    
  write.csv(as.data.frame(unique_infected_compare_KEGG),
            file = file.path(charts_folder,
                             "unique_infected_compare_Kegg.csv"),
            row.names = FALSE)
  unique_infected_compare <- unique_infected_compare_KEGG
}
```

# Save results
```{r}
saveRDS(unique_infected_compare_GO,
        file.path(GO_save_folder,"unique_infected_compare_GO.Rds"))
saveRDS(unique_infected_compare_KEGG,
        file.path(pathway_save_folder,"unique_infected_compare_kegg.Rds"))
saveRDS(unique_infected_compare_reactome,
        file.path(pathway_save_folder,"unique_infected_compare_reactome.Rds"))
```

```{r}
unique_infected_compare_GO <-
  readRDS(file.path(GO_save_folder,"unique_infected_compare_GO.Rds"))
```

```{r}
generate_volcano_plots(results_list = results.list[1:4],
                       pos_color = "#E41A1C",
                       neg_color = "#377EB8",
                       plots_folder = plots_folder,
                       charts_folder = charts_folder
                       )
```

```{r}
generate_volcano_plots_mark_unique(
  results_list = results.list,
  unique_genes_list = unique_genes_list,
  pos_color = "#E41A1C",
  neg_color = "#377EB8",
  mark_color = "purple",
  plots_folder = plots_folder,
  charts_folder = charts_folder,
  filename_prefix = "Full_unique_purple",
  save = TRUE,
  subset_condition = 1:4
  )
```


```{r}
volcanos_unique_infected <- generate_volcano_plots_mark_unique(
  results_list = results.list[1:4],
  unique_genes_list = unique_genes_list,
  pos_color = "#E41A1C",
  neg_color = "#377EB8",
  mark_color = "purple",
  plots_folder = plots_folder,
  charts_folder = charts_folder,
  filename_prefix = "Full_unique_purple",
  save = FALSE
  )
```
