---
title: "tf_analysis_modified (version 2025)"
author: "Johannes Schöneich"
date: '`r Sys.Date()`'
output: html_document
---

# Libraries
```{r setup, include = FALSE}
## We load the required packages
library(LR2TF)
library(tidyverse)
library(ComplexHeatmap)
library(grid)
library(stringr)
library(maditr)
library(circlize)
library(glue)
library(circlize)
library(ComplexHeatmap)
library(Seurat)
library(cowplot)

harddrive <- "/media/johannes/EXTERNAL_USB/231248/NTFS_3.64TB"

source(file.path(harddrive,"helper_functions.R"))
source(file.path(harddrive,"scrna_datasets/save_load_helper.R"))
set.seed(1234)
```

# Adjustments
```{r}
condition <- "prWe"

analysis_folder <- file.path(harddrive,
                          "dorothea_tf_analysis/analysis/TF_results")

condition_folder <- file.path(analysis_folder,condition)

comparison_df <- readRDS(file.path(analysis_folder,
                                   "compared/comparison_dfs.Rds")
                         )

names(comparison_df) <- make.names(names(comparison_df))

tf_vec <- c("Pdx1","Gata5","Hnf1a","Cdx2","Cdp","Sox9","Sox7","Sox17",
            "Sox18","Hoxa13","Hoxd13","Spdef","Hes1","Atoh1","Wnt3a",
            "Notch1","Ngn3","Tcf7l1","Klf4","Math1","Foxa2", "Gata4",
            "Fgf4","Gfi1","Foxf1","Foxl1","Ctnnb1","Prdm1","Mafb","Spib",
            "Tcf7l2")

plots_folder <- file.path(analysis_folder,"plots")
charts_folder <- file.path(analysis_folder,"charts")

dir.create(plots_folder)
dir.create(charts_folder)
```

```{r}
comparison_df$prWe.vs.poWe$CellType <- gsub("+",".",
                      comparison_df$prWe.vs.poWe$CellType,
                      fixed = TRUE)

rownames(comparison_df$prWe.vs.poWe) <- gsub("+",".",
                       rownames(comparison_df$prWe.vs.poWe),
                       fixed = TRUE)

tf_activity_tables <- comparison_df

tf_activity_tables$prWe.vs.poWe <- tf_activity_tables$prWe.vs.poWe %>% 
 filter(tf %in% tf_vec)
```

# Calculate tf heatmap
```{r}
calculate_tf_heatmap <- function(nm_df, title, top_bot = FALSE){
  nm_df <- as.data.frame(nm_df)
  significant_res <- nm_df[nm_df$tag == "***", ]
  significant_genes <- unique(significant_res$tf)
  nm_df <- filter(nm_df, nm_df$tf %in% significant_genes)

  
  nm_df_short <- nm_df[c("r", "tf", "CellType")]
  nm_df_clust <- tapply(nm_df_short$r, list(tf = nm_df_short$tf, 
                        CellType = nm_df_short$CellType),
                        mean)
  
  #nm_df_clust <- data.frame(nm_df_clust)
  #nm_df_clust <- nm_df_clust[rownames(nm_df_clust) %in% tf_vec,] 
  nm_df_clust <- as.matrix(nm_df_clust)
  
  if(top_bot == TRUE){
   nm_df_clust <- as.data.frame(nm_df_clust) 
   nm_df_clust$gene <- rownames(nm_df_clust)
 
   df_long <- nm_df_clust %>%
     pivot_longer(cols = c(Abs.Ent., EEC, Goblet.Paneth), 
               names_to = "zscore_type", values_to = "zscore")
 
   bot3 <- df_long %>% arrange(zscore) %>% group_by(zscore_type) %>% 
     slice_head(n = 3)
   top3 <- df_long %>% arrange(zscore) %>% group_by(zscore_type) %>% 
     slice_tail(n = 3)
   nm_df_clust <- nm_df_clust %>% filter(nm_df_clust$gene %in% 
                                           c(bot3$gene,top3$gene)) %>% 
     select(!gene) %>% as.matrix()
  }
  
  tag_mapping <- nm_df[c("tf", "tag", "CellType")]
  tag_mapping <- dcast(tag_mapping, tf ~ CellType, value.var = "tag")

  #tag_mapping <- tag_mapping[rownames(tag_mapping) %in% tf_vec,] 
  tag_mapping <- as.data.frame(tag_mapping)
  
  rownames(tag_mapping) <- tag_mapping$tf
  tag_mapping$tf <- NULL
  colnames(tag_mapping) <- colnames(nm_df_clust)
   
  col_min <- round(min(nm_df_clust, na.rm = TRUE),digits = 1)
  col_max <- round(max(nm_df_clust, na.rm = TRUE),digits = 1)

  fh <- function(x) fastcluster::hclust(dist(x))
  p <- Heatmap(nm_df_clust, 
        name = "r", 
        cluster_rows = fh,
        cluster_columns = FALSE, 
        width = ncol(nm_df_clust) * unit(4, "mm"),
        height = nrow(nm_df_clust) * unit(4, "mm"),
        row_title = "Transcription Factor", 
        column_title = title,
        column_title_gp = gpar(fontsize = 7),
        col = colorRamp2(c(col_min, 0, col_max), 
                c("blue", "white", "red")),
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 7),
        column_labels = gsub("Goblet.Paneth","Goblet+Paneth",
                             colnames(nm_df_clust),fixed = TRUE),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(tag_mapping[as.character(rownames(nm_df_clust)[i]),
                    as.character(colnames(nm_df_clust)[j])],
                    x, y,
                    gp = gpar(fontsize = 7))
         })
  return(p)
}

#function for condition
calculate_tf_condition_heatmap <- function(seuratobject.markers, viper_scores, 
                                           title, top_bot = FALSE){
  seuratobject.markers$tag <- sapply(seuratobject.markers$p_val_adj,
                   function(pval) {
  if (pval < 0.001) {
   txt <- "***"
  } else if (pval < 0.01) {
   txt <- "**"
  } else if (pval < 0.05) {
   txt <- "*"
  } else {
   txt <- "ns"
  }
  return(txt)
 })
 
 seuratobject.markers$log_fc_tag <- sapply(seuratobject.markers$avg_log2FC, 
                      function(log_fc) {
  if (log_fc >= 1.0) {
   txt <- "***"
  } else if (log_fc > 0.5) {
   txt <- "**"
  } else if (log_fc > 0.0) {
   txt <- "*"
  } else {
   txt <- "ns"
  }
  return(txt)
 })
 
 tag_mapping <- seuratobject.markers[c("gene", "tag", "log_fc_tag", 
                    "cluster", "p_val_adj", "avg_log2FC")]
 
 tag_mapping <- as.data.frame(dcast(tag_mapping, gene ~ cluster, value.var = "tag"))
 rownames(tag_mapping) <- tag_mapping$gene
 
 tag_mapping$gene <- NULL
 tag_mapping[is.na(tag_mapping)] <- "ns"
 
 tag_mapping <- tag_mapping[rownames(viper_scores),]
 colnames(tag_mapping) <- gsub("+",".",colnames(tag_mapping),fixed = TRUE)
 
 tf_scores <- as.matrix(viper_scores)

 if(top_bot == TRUE){
   tf_scores <- as.data.frame(tf_scores) 
   tf_scores$gene <- rownames(tf_scores)
 
   df_long <- tf_scores %>%
     pivot_longer(cols = c(Abs.Ent., EEC, Goblet.Paneth, Stem), 
               names_to = "zscore_type", values_to = "zscore")
 
   bot3 <- df_long %>% arrange(zscore) %>% group_by(zscore_type) %>% slice_head(n = 3)
   top3 <- df_long %>% arrange(zscore) %>% group_by(zscore_type) %>% slice_tail(n = 3)
   tf_scores <- tf_scores %>% filter(tf_scores$gene %in% c(bot3$gene,top3$gene)) %>% 
     select(!gene) %>% as.matrix()
 }
 
 column_names_table <- colnames(tf_scores)
 column_names_tags <- colnames(tag_mapping)
 
 if (length(column_names_tags) < length(column_names_table)) {
  missing_col <- setdiff(column_names_table, column_names_tags)
  for (col in missing_col) {
   tag_mapping[col] <- "ns"
  }
 }
 
 rownames(tf_scores) <- gsub(".", "-", rownames(tf_scores), fixed = TRUE)
 
 #col_min <- round(min(tf_scores, na.rm = TRUE),digits = 1)
 #col_max <- round(max(tf_scores, na.rm = TRUE),digits = 1)
 
 col_min = -4
 col_max = 4
 
 fh <- function(x) fastcluster::hclust(dist(x))
 p <- Heatmap(tf_scores,
        name = "z-score",
        #cluster_rows = FALSE,
        cluster_rows = fh,
        cluster_columns = FALSE,
        width = ncol(tf_scores) * unit(4, "mm"),
        height = nrow(tf_scores) * unit(4, "mm"),
        #row_title = "Transcription Factor", 
        #column_title = glue("Celltype ({title})"),
        column_title = title,
        column_labels = gsub("Goblet.Paneth","Goblet+Paneth",
                             colnames(tf_scores),fixed = TRUE),
        row_names_gp = gpar(fontsize = 7), 
        row_names_rot = 45,
        column_names_gp = gpar(fontsize = 7),
        heatmap_legend_param = c(labels_gp = gpar(fontsize = 7),
                                                  title_gp = gpar(fontsize = 7, 
                                                                  fontface = "bold")),
        col = colorRamp2(c(col_min, 0, col_max), c("blue", "white", "red")),
        cell_fun = function(j, i, x, y, width, height, fill) {
         grid.text(tag_mapping[as.character(rownames(tf_scores)[i]), 
                    as.character(colnames(tf_scores)[j])],
              x, y, 
              gp = gpar(fontsize = 7))
        })
 return(p)
}
```


```{r}
calculate_tf_condition_heatmap_transposed <- function(seuratobject.markers, 
                                                      viper_scores,
                                                      title, 
                                                      top_bot = FALSE) {
  
  seuratobject.markers$tag <- sapply(seuratobject.markers$p_val_adj,
                   function(pval) {
    if (pval < 0.001) {
      txt <- "***"
    } else if (pval < 0.01) {
      txt <- "**"
    } else if (pval < 0.05) {
      txt <- "*"
    } else {
      txt <- "ns"
    }
    return(txt)
  }
  )
 
  seuratobject.markers$log_fc_tag <- sapply(seuratobject.markers$avg_log2FC, 
                      function(log_fc) {
    if (log_fc >= 1.0) {
      txt <- "***"
    } else if (log_fc > 0.5) {
      txt <- "**"
    } else if (log_fc > 0.0) {
      txt <- "*"
    } else {
      txt <- "ns"
    }
    return(txt)
  }
  )
 
  tag_mapping <- seuratobject.markers[c("gene", "tag", "log_fc_tag", 
                    "cluster", "p_val_adj", "avg_log2FC")]
 
  tag_mapping <- as.data.frame(dcast(tag_mapping, gene ~ cluster, value.var = "tag"))
  rownames(tag_mapping) <- tag_mapping$gene
 
  tag_mapping$gene <- NULL
  tag_mapping[is.na(tag_mapping)] <- "ns"
 
  tag_mapping <- tag_mapping[rownames(viper_scores),]
  colnames(tag_mapping) <- gsub("+", ".", colnames(tag_mapping), fixed = TRUE)
 
  tf_scores <- as.matrix(viper_scores)

  if (top_bot == TRUE) {
    tf_scores <- as.data.frame(tf_scores) 
    tf_scores$gene <- rownames(tf_scores)
 
    df_long <- tf_scores %>%
      pivot_longer(cols = c(Abs.Ent., EEC, Goblet.Paneth, Stem), 
               names_to = "zscore_type", values_to = "zscore")
 
    bot3 <- df_long %>% arrange(zscore) %>% group_by(zscore_type) %>% slice_head(n = 3)
    top3 <- df_long %>% arrange(zscore) %>% group_by(zscore_type) %>% slice_tail(n = 3)
    tf_scores <- tf_scores %>% filter(tf_scores$gene %in% c(bot3$gene, top3$gene)) %>% 
      select(!gene) %>% as.matrix()
  }
 
  column_names_table <- colnames(tf_scores)
  column_names_tags <- colnames(tag_mapping)
 
  if (length(column_names_tags) < length(column_names_table)) {
    missing_col <- setdiff(column_names_table, column_names_tags)
    for (col in missing_col) {
      tag_mapping[col] <- "ns"
    }
  }
 
  rownames(tf_scores) <- gsub(".", "-", rownames(tf_scores), fixed = TRUE)
 
  # Transpose matrices for swapped axes
  tf_scores <- t(tf_scores)
  tag_mapping <- t(tag_mapping)
 
  col_min <- round(min(tf_scores, na.rm = TRUE), digits = 1)
  col_max <- round(max(tf_scores, na.rm = TRUE), digits = 1)
 
  fh <- function(x) fastcluster::hclust(dist(x))
  
  p <- Heatmap(
    tf_scores,
    name = "z-score",
    cluster_rows = FALSE,
    cluster_columns = fh,
    width = ncol(tf_scores) * unit(4, "mm"),
    height = nrow(tf_scores) * unit(4, "mm"),
    #row_title = "Cell Types/Clusters", 
    #column_title = glue("Genes ({title})"),
    column_labels = colnames(tf_scores),  # Updated to match transposed matrix
    column_names_gp = gpar(fontsize = 7),
    row_names_gp = gpar(fontsize = 7),
    row_names_side = "left",
    column_names_rot = 45,
    heatmap_legend_param = list(
      labels_gp = gpar(fontsize = 7),
      title_gp = gpar(fontsize = 7, fontface = "bold")
    ),
    col = colorRamp2(c(col_min, 0, col_max), c("blue", "white", "red")),
    cell_fun = function(j ,i, x, y, w, h, fill) {
      grid.text(
        tag_mapping[i, j],
        x = x,
        y = y,
        gp = gpar(fontsize = 7)
      )
    }
  )
  return(p)
}


calc_ht_size <- function(ht, unit = "inch"){
    pdf(NULL)
    ht <- draw(ht)
    
    # Calculate the width and height of the heatmap
    heatmap_width <- ComplexHeatmap:::width(ht)
    heatmap_width <- convertX(heatmap_width, unit, 
                              valueOnly = TRUE)
    heatmap_height <- ComplexHeatmap:::height(ht)
    heatmap_height <- convertY(heatmap_height, unit, 
                               valueOnly = TRUE)
    dev.off()
    
    fs <- ht@column_title_param$gp$fontsize
    
    title_char <- ht@column_title
    if(is_empty(title_char)){
      title_char <- ht@ht_list[[1]]@column_title
    }
    
    # font does not use font size,but something else, so we need to check with par.
    # implement when necessary
    col_title_width <- strwidth(title_char, units = "inch", 
                                font = fs)
    
    # Use the maximum width between the heatmap and the column title
    total_width <- max(heatmap_width, col_title_width)
    
    return(c(total_width, heatmap_height))
}
```

```{r}
hm <- calculate_tf_heatmap(tf_activity_tables$prWe.vs.poWe,
                           #title = "prWe.vs.poWe r effect size"
                           title = element_blank())

size <- calc_ht_size(hm)
```

```{r}
save_ggplot_formats(
  plt = hm,
  base_plot_dir = plots_folder,
  plt_name = "cluster_condition_activity_difference_v3",
  width = size[1], height = size[2],
  plot_obj = "heatmap"
)
```

## use subset top 5 and subset top 3, bottom 3 (major changes) per cell type
```{r}
df <- comparison_df$prWe.vs.poWe

### top 5 subset
top5_vec <- df %>% arrange(FDR) %>% group_by(CellType) %>%
  slice_head(n = 5) %>% pull(tf) %>% unique()

top5df <- comparison_df$prWe.vs.poWe %>% 
 filter(tf %in% top5_vec)

hm_top5 <- calculate_tf_heatmap(top5df,
  title = "prWe.vs.poWe r effect size (top 5 per celltype)"
  )

size <- calc_ht_size(hm_top5)
```

```{r}
save_ggplot_formats(
  plt = hm_top5,
  base_plot_dir = plots_folder,
  plt_name = "cluster_condition_activity_difference_top5_v3",
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```

### top bottom 3
```{r}
hm_topbot <- calculate_tf_heatmap(df,
    title = "prWe.vs.poWe r effect size (top/bottom 3 per celltype)",
    top_bot = TRUE)

size <- calc_ht_size(hm_topbot)

save_ggplot_formats(
  plt = hm_topbot,
  base_plot_dir = plots_folder,
  plt_name = "cluster_condition_activity_difference_topbot_v3",
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```

# Calculate transcription factor heatmap inside condition 
```{r}
condition <- "prWe"

analysis_folder <- file.path(harddrive,
                             "dorothea_tf_analysis/analysis/TF_results")

condition_folder <- file.path(analysis_folder,condition)

seuratobject.markers <- read.csv(
  file.path(condition_folder,
            glue("all_specificmarker__{condition}.csv")
            )
  )

viper_scores <- read.csv(file.path(condition_folder,
                                   glue("tf_scores_{condition}.csv")),
                         row.names = 1
        )

viper_scores_filtered <- viper_scores[rownames(viper_scores) %in% tf_vec,]
```

```{r}
condition_hm <- calculate_tf_condition_heatmap(
  seuratobject.markers = seuratobject.markers,
  viper_scores = viper_scores,
  #title = condition
  title = element_blank()
  )

size <- calc_ht_size(condition_hm)
```


```{r}
condition_hm_tp <- calculate_tf_condition_heatmap_transposed(
  seuratobject.markers = seuratobject.markers,
  viper_scores = viper_scores,
  #title = condition
  title = element_blank()
  )

size2 <- calc_ht_size(condition_hm_tp)

save_ggplot_formats(
  plt = condition_hm_tp,
  base_plot_dir = plots_folder,
  plt_name = glue("tf_activity_{condition}_test"),
  width = size2[1], height = size2[2],
  plot_obj = "heatmap"
  )
```

```{r}
save_ggplot_formats(
  plt = condition_hm,
  base_plot_dir = plots_folder,
  plt_name = glue("tf_activity_{condition}_v3"),
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```

# condition subset top 5 and top bottom 3
```{r}
seurat.markers_top5 <- seuratobject.markers %>% arrange(p_val_adj) %>%
  group_by(cluster) %>% slice_head(n = 5)

viper_scores_top5 <- viper_scores %>% 
  filter(rownames(viper_scores) %in% seurat.markers_top5$gene)

condition_hm_top5 <- calculate_tf_condition_heatmap(
  seuratobject.markers = seurat.markers_top5,
  viper_scores = viper_scores_top5,
  #title = condition
  title = element_blank()
  )

size <- calc_ht_size(condition_hm_top5)
```

```{r}
save_ggplot_formats(
  plt = condition_hm_top5,
  base_plot_dir = plots_folder,
  plt_name = glue("tf_activity_top5_{condition}_v3"),
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```

```{r}
condition_hm_topbot <- calculate_tf_condition_heatmap(
    seuratobject.markers = seuratobject.markers,
    viper_scores = viper_scores,
    title = condition,
    top_bot = TRUE
    )

size <- calc_ht_size(condition_hm_topbot)

save_ggplot_formats(
  plt = condition_hm_topbot,
  base_plot_dir = plots_folder,
  plt_name = glue("tf_activity_topbot_{condition}_v3"),
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```

# Real transcription factor expression
```{r}
#transcription factor expression in heatmap (similar to the ones we already produced)

scrna_folder <- file.path(harddrive,
                          "scrna_datasets/scrna_both_d25/scrna_full_2024_both_d25")

cluster_use <- "int_0.3_broad"

scrna <- load_object(
  file.path(scrna_folder,"scrna_uninf_bothd25.Rds")
  )

DefaultAssay(scrna) <- "RNA"

Idents(scrna) <- cluster_use

scrna_filtered <- subset(scrna, idents = c("Tuft"), invert = TRUE)
```

```{r}
# Get the genenames in the row order f
drawn_hm <- draw(hm_top5)
row_order <- drawn_hm@ht_list$r@row_order

genename <- rownames(hm_top5@matrix)[row_order]
genename <- genename[genename %in% rownames(scrna_filtered@assays$RNA)]

# Assuming scrna is your Seurat object and genename is your list of genes
# Calculate average expression for each cluster
average_expression <- AverageExpression(scrna_filtered, features = genename,
                                        return.seurat = TRUE,
                                        assays = "RNA")
average_expression_matrix <- GetAssayData(average_expression, slot = "data")

# Initialize matrices to store p-values and adjusted p-values (FDR)
p_value_matrix <- matrix(nrow = length(genename), 
                         ncol = length(levels(scrna_filtered))
                         )

rownames(p_value_matrix) <- genename
colnames(p_value_matrix) <- levels(scrna_filtered)

fdr_matrix <- p_value_matrix

# Perform differential expression analysis for each cluster
for (cluster in levels(scrna_filtered)) {
  prWe_cells <- WhichCells(scrna_filtered, idents = cluster, 
                           expression = stage == 'prWe')
  poWe_cells <- WhichCells(scrna_filtered, idents = cluster,
                           expression = stage == 'poWe')
  
  # Ensure there are enough cells for comparison
  if (length(prWe_cells) > 1 & length(poWe_cells) > 1) {  
    markers <- FindMarkers(scrna_filtered, ident.1 = prWe_cells,
                           ident.2 = poWe_cells, features = genename)
    
    # Match gene names and fill in p_value_matrix and fdr_matrix
    matched_genes <- intersect(genename, rownames(markers))
    p_value_matrix[matched_genes, cluster] <- markers[matched_genes, "p_val"]
    fdr_matrix[matched_genes, cluster] <- p.adjust(markers[matched_genes, "p_val"],
                                                   method = "fdr")
  } else {
    p_value_matrix[, cluster] <- NA
    fdr_matrix[, cluster] <- NA
  }
}
```

```{r}
# Function to calculate effect size
effect_size <- function(x, y) {
  n_x <- length(x)
  n_y <- length(y)
  mean_diff <- mean(x) - mean(y)
  pooled_sd <- sqrt(((n_x - 1) * stats::var(x) + (n_y - 1) * stats::var(y)) / (n_x + n_y - 2))
  r <- mean_diff / pooled_sd
  return(r)
}

# Initialize a matrix to store effect sizes
effect_size_matrix <- matrix(nrow = length(genename), 
                             ncol = length(levels(scrna_filtered))
                             )
rownames(effect_size_matrix) <- genename
colnames(effect_size_matrix) <- levels(scrna_filtered)

# Calculate effect sizes for each gene in each cluster
for (cluster in levels(scrna_filtered)) {
  prWe_cells <- WhichCells(scrna_filtered, idents = cluster, 
                           expression = stage == 'prWe')
  poWe_cells <- WhichCells(scrna_filtered, idents = cluster, 
                           expression = stage == 'poWe')
  
  for (gene in genename) {
    prWe_data <- FetchData(scrna_filtered, vars = gene, cells = prWe_cells)[[gene]]
    poWe_data <- FetchData(scrna_filtered, vars = gene, cells = poWe_cells)[[gene]]
    
    if (length(prWe_data) > 1 & length(poWe_data) > 1) {
      effect_size_matrix[gene, cluster] <- effect_size(prWe_data, poWe_data)
    } else {
      effect_size_matrix[gene, cluster] <- NA
    }
  }
}
# Function to convert p-value to significance stars
get_significance_stars <- function(pval) {
  if (is.na(pval)) {
    return("")
  } else if (pval < 0.001) {
    return("***")
  } else if (pval < 0.01) {
    return("**")
  } else if (pval < 0.05) {
    return("*")
  } else {
    return("ns")
  }
}

col_min <- round(min(effect_size_matrix, na.rm = TRUE),digits = 1)
col_max <- round(max(effect_size_matrix, na.rm = TRUE),digits = 1)
```

```{r}
# Generate heatmap
hm_effectsize <- Heatmap(as.matrix(effect_size_matrix), 
        name = "r", 
        #cluster_rows = fh,
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        width = ncol(effect_size_matrix) * unit(4, "mm"),
        height = nrow(effect_size_matrix) * unit(4, "mm"), 
        row_title = "Transcription Factor",
        #column_title = "Transcription factor expression change in single cell",
        col = colorRamp2(c(col_min, 0, col_max), 
                         c("blue", "white", "red")),
          row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 7),
        cell_fun = function(j, i, x, y, width, height, fill) {
          sig_stars <- get_significance_stars(effect_size_matrix[i, j])
          grid.text(sig_stars, x, y, gp = gpar(fontsize = 7))
        }
        )

size <- calc_ht_size(hm_effectsize)
```

```{r}
save_ggplot_formats(
  plt = hm_effectsize,
  base_plot_dir = plots_folder,
  plt_name = glue("tf_expression_in_sc_effect_size_prWe_poWe_v2"),
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```

## Only the top 5 pre Weaning
```{r}
scrna_prWe <- subset(scrna_filtered, subset = stage == 'prWe')
```

```{r}
drawn_hm <- draw(condition_hm_top5)
row_order <- drawn_hm@ht_list[[1]]@row_order

genename <- rownames(condition_hm_top5@matrix)[row_order]

#Find missing genes
missing_genes <- setdiff(genename,rownames(scrna_prWe@assays$RNA))

intersect_genes <- intersect(genename,rownames(scrna_prWe@assays$RNA))

#Objective: Add missing genes on the exact same position as before, also do this in the pvalue matrix

# Calculate average expression for each cluster
average_expression_prWe <- AverageExpression(scrna_prWe, 
                                             features = intersect_genes,
                                             return.seurat = TRUE,
                                             assays = "RNA")
average_expression_matrix_prWe <- GetAssayData(average_expression_prWe, slot = "data")
```

```{r}
# Standardize (z-score) the average expression values
z_score_matrix <- t(scale(t(average_expression_matrix_prWe)))

# Function to calculate p-values for differential expression between clusters
calculate_p_values <- function(scrna_obj, genes) {
  clusters <- levels(scrna_obj)
  p_value_matrix <- matrix(NA, nrow = length(genes), ncol = length(clusters))
  rownames(p_value_matrix) <- genes
  colnames(p_value_matrix) <- clusters
  
  for (gene in genes) {
    expr_values <- FetchData(scrna_obj, vars = gene)
    p_vals <- sapply(clusters, function(cluster) {
      cluster_cells <- WhichCells(scrna_obj, idents = cluster)
      other_cells <- setdiff(Cells(scrna_obj), cluster_cells)
      t.test(expr_values[cluster_cells,], expr_values[other_cells,])$p.value
    })
    p_value_matrix[gene, ] <- p.adjust(p_vals, method = "fdr")
  }
  return(p_value_matrix)
}

# Calculate p-values for the prWe cells
p_value_matrix_prWe <- calculate_p_values(scrna_prWe, intersect_genes)

col_min <- round(min(z_score_matrix, na.rm = TRUE),digits = 1)
col_max <- round(max(z_score_matrix, na.rm = TRUE),digits = 1)
```

```{r}
add_missing_genes <- function(df, missing_genes, gene_vector) {
  df <- as.data.frame(df)
  # Create a dataframe for missing genes with expression levels set to NA
  num_cols <- ncol(df)
  missing_df <- data.frame(matrix(NA, nrow = length(missing_genes), ncol = num_cols))
  rownames(missing_df) <- missing_genes
  colnames(missing_df) <- colnames(df)
  
  # Add missing genes to the original dataframe
  df <- rbind(df, missing_df)
  
  # Sort the dataframe to match the order of `gene_vector`
  df <- df[match(gene_vector, rownames(df)), ]
  
  return(df)
}

z_score_matrix_mg <- add_missing_genes(z_score_matrix,missing_genes,genename)

p_value_matrix_prWe_mg <- add_missing_genes(p_value_matrix_prWe,missing_genes,genename)

z_score_matrix_mg_transposed <- t(z_score_matrix_mg)
p_value_matrix_prWe_mg_transposed <- t(p_value_matrix_prWe_mg)


# colnames(z_score_matrix_mg) <- gsub("+",".",
#                        colnames(z_score_matrix_mg),
#                        fixed = TRUE)
# 
# colnames(p_value_matrix_prWe_mg) <- gsub("+",".",
#                        colnames(p_value_matrix_prWe_mg),
#                        fixed = TRUE)

```

```{r}
col_min = -4
col_max = 4
# Generate heatmap
hm_prWe_same_maxima <- Heatmap(as.matrix(z_score_matrix_mg), 
        name = "z-score", 
        #cluster_rows = fh,
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        width = ncol(z_score_matrix_mg) * unit(4, "mm"),
        height = nrow(z_score_matrix_mg) * unit(4, "mm"), 
        #row_title = "Transcription Factor",
        #column_title = "Clusterwise tf expression change in single cell (prWe)",
        column_labels = colnames(z_score_matrix_mg),
        col = colorRamp2(c(col_min, 0, col_max), 
                         c("blue", "white", "red")),
        row_names_gp = gpar(fontsize = 7),
        row_names_rot = 45,
        column_names_gp = gpar(fontsize = 7),
        cell_fun = function(j, i, x, y, width, height, fill) {
          sig_stars <- get_significance_stars(p_value_matrix_prWe_mg[i, j])
          grid.text(sig_stars, x, y, gp = gpar(fontsize = 7))
        }
)

size <- calc_ht_size(hm_prWe_same_maxima)
```

```{r}
col_min = -4
col_max = 4
# Generate heatmap
hm_prWe_same_maxima_transposed <- Heatmap(as.matrix(z_score_matrix_mg_transposed), 
        name = "z-score", 
        #cluster_rows = fh,
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        width = ncol(z_score_matrix_mg_transposed) * unit(4, "mm"),
        height = nrow(z_score_matrix_mg_transposed) * unit(4, "mm"), 
        #row_title = "Transcription Factor",
        #column_title = "Clusterwise tf expression change in single cell (prWe)",
        row_labels = rownames(z_score_matrix_mg_transposed),
        col = colorRamp2(c(col_min, 0, col_max), 
                         c("blue", "white", "red")),
        row_names_gp = gpar(fontsize = 7),
        row_names_side = "left",
        column_names_rot = 45,
        column_names_gp = gpar(fontsize = 7),
        cell_fun = function(j, i, x, y, width, height, fill) {
          sig_stars <- get_significance_stars(p_value_matrix_prWe_mg_transposed[i, j])
          grid.text(sig_stars, x, y, gp = gpar(fontsize = 7))
        }
)

size <- calc_ht_size(hm_prWe_same_maxima_transposed)
```



```{r}
col_min <- round(min(z_score_matrix, na.rm = TRUE),digits = 1)
col_max <- round(max(z_score_matrix, na.rm = TRUE),digits = 1)

# Generate heatmap
hm_prWe <- Heatmap(as.matrix(z_score_matrix_mg), 
        name = "z-score", 
        #cluster_rows = fh,
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        width = ncol(z_score_matrix_mg) * unit(4, "mm"),
        height = nrow(z_score_matrix_mg) * unit(4, "mm"), 
        #row_title = "Transcription Factor",
        #column_title = "Clusterwise tf expression change in single cell (prWe)",
        column_labels = colnames(z_score_matrix_mg),
        col = colorRamp2(c(col_min, 0, col_max), 
                         c("blue", "white", "red")),
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 7),
        cell_fun = function(j, i, x, y, width, height, fill) {
          sig_stars <- get_significance_stars(p_value_matrix_prWe_mg[i, j])
          grid.text(sig_stars, x, y, gp = gpar(fontsize = 7))
        }
)

size <- calc_ht_size(hm_prWe)
```

```{r}
save_ggplot_formats(
  plt = hm_prWe,
  base_plot_dir = plots_folder,
  plt_name = glue("tf_expression_in_sc_prWe_v3"),
  width = size[1], height = size[2],
  plot_obj = "heatmap"
  )
```


# Calculate again to make them all in one plot together -----
```{r}
hm <- calculate_tf_heatmap(tf_activity_tables$prWe.vs.poWe,
                           #title = "prWe.vs.poWe r effect size"
                           title = element_blank())

condition_hm_top5 <- calculate_tf_condition_heatmap(
  seuratobject.markers = seurat.markers_top5,
  viper_scores = viper_scores_top5,
  #title = condition
  title = element_blank()
  )
```

#do transposed heatmap for the paper
```{r}
condition_hm_top5_v2 <- calculate_tf_condition_heatmap_transposed(
  seuratobject.markers = seurat.markers_top5,
  viper_scores = viper_scores_top5,
  #title = condition
  title = element_blank()
  )
```

```{r}
col_min = -0.6
col_max = 0.3

# Create text-only legends for annotations
top_annotation <- Legend(
  at = character(0),
  background = NA,
  border = NA,
  labels = "prWe", 
  title = NULL, 
  type = "text",
  labels_gp = gpar(fontsize = 7)
)

bottom_annotation <- Legend(
  at = character(0),
  background = NA,
  border = NA,
  labels = "poWe", 
  title = NULL, 
  type = "text",
  labels_gp = gpar(fontsize = 7)
)

# Create the main heatmap legend
heatmap_legend <- Legend(
  col_fun = colorRamp2(c(col_min, 0, col_max), c("blue", "white", "red")), 
  title = "r", 
  title_gp = gpar(fontsize = 7, fontface = "bold"),
  labels_gp = gpar(fontsize = 7)
)

# Combine the legends into a single legend object
combined_legend_top_mid <- packLegend(
  top_annotation, 
  heatmap_legend, 
  direction = "vertical",
  gap = unit(-2,"mm")
)

combined_legend <- packLegend(
  combined_legend_top_mid,
  bottom_annotation,
  direction = "vertical",
  gap = unit(2,"mm")
)

# Draw the heatmap with the combined legend next to it
draw(hm, show_heatmap_legend = FALSE)
draw(combined_legend, x = unit(0.73, "npc"), 
     y = unit(0.65, "npc"), just = c("right", "center"))
```

```{r}
hm_grid <- grid.grabExpr({draw(hm, show_heatmap_legend = FALSE, background = "transparent")
                          draw(combined_legend, x = unit(0.89, "npc"), 
                               y = unit(0.52, "npc"), just = c("right", "center")
                          )
                          }
     )

condition_hm_grid_no_legend <- condition_hm_top5 %>% draw(show_heatmap_legend = FALSE,
                                                          background = "transparent") %>% 
        grid.grabExpr()

condition_hm_grid_legend <- condition_hm_top5 %>% draw(show_heatmap_legend = TRUE,
                                                       background = "transparent") %>% 
        grid.grabExpr()

hm_prWe_grid_sm <- hm_prWe_same_maxima %>% draw(background = "transparent") %>% 
        grid.grabExpr()

hm_prWe_grid <- hm_prWe %>% draw(background = "transparent") %>% 
        grid.grabExpr()
```

```{r}
heatmap_grid_two_legend <- plot_grid(condition_hm_grid_no_legend,
                                     hm_prWe_grid_sm,
                                     rel_widths = c(1, 1),
                                     nrow = 1)
save_ggplot_formats(
  plt = heatmap_grid_two_legend,
  base_plot_dir = plots_folder,
  plt_name = glue("heatmap_grid_two_legend_v4"),
  width = 4, height = 6
  )
```

#Transposed Heatmaps

```{r}
condition_hm_grid_no_legend_transposed <- condition_hm_top5_v2 %>% 
  draw(show_heatmap_legend = FALSE, background = "transparent") %>% 
        grid.grabExpr()

hm_prWe_grid_sm_transposed <- hm_prWe_same_maxima_transposed %>% 
  draw(show_heatmap_legend = FALSE,background = "transparent") %>% 
        grid.grabExpr()
```

```{r}
heatmap_grid_transposed <- plot_grid(condition_hm_grid_no_legend_transposed,
                                     hm_prWe_grid_sm_transposed,
                                     rel_widths = c(1, 1),
                                     nrow = 2)
save_ggplot_formats(
  plt = heatmap_grid_transposed,
  base_plot_dir = plots_folder,
  plt_name = glue("heatmap_grid_two_legend_v5"),
  width = 4, height = 2.5
  )
```


```{r}
#save workspace
save.image(file.path(harddrive,"workspaces/workspace_tf_analysis.Rdata"))
```
